% Acknowledgements ============================================================

\pdfbookmark[1]{Acknowledgments}{acknowledgments}
\chapter*{Acknowledgments}
I would like to express my sincere gratitude to the members of my jury, Alexandros Kalousis, J{\"o}rn Jacobsen, Patrick Gallinari, Pierre Sacr{\'e}, and Pierre Geurts. I greatly appreciated their insightful comments, constructive criticism, and the engaging discussions we had during the public defence.

I feel lucky to have had the opportunity to learn from Pierre during his data structure and algorithms course. His teaching ignited my passion for computer science. I am grateful for the foundation he provided that enabled me to pursue my research interests.

I want to give a special thanks to J{\"o}rn and Jens Behrmann for the chance they gave me to work with them as an intern at Apple. Their positivity, expertise, and challenging approaches significantly impacted my personal and professional growth. I feel fortunate to have worked with such exceptional researchers and mentors.

I am incredibly fortunate to have had Gilles Louppe as my advisor throughout my doctoral studies. Gilles, you have been an outstanding advisor, constantly challenging my assumptions and supporting me in my desire to explore new ideas. I feel grateful for your unwavering support, encouragement, and guidance. The many coffee breaks we shared were not only enjoyable but also helped me to grow new perspectives and insights on what it is to be a researcher; I will certainly miss them.

I would like to thank my colleagues at the department for their camaraderie, support, and the great memories we shared. Nicolas, my office mate for over three years, you made my workdays awesome, even though the quality of your humour is debatable. I feel honoured that you trusted me to be the godfather of Elena. Pascal, who always brings good vibes to wherever you go, I was lucky to share four years of PhD with you. Nicolas, and Pascal, we should also be grateful to Van Anh for being so kind to tolerate almost all of our terrible jokes. Damien, Nicolas, Pascal, and Gilles, thanks for the tremendous Orval-intensive times we had after work, even though it was sometimes a bit harsh the day after. I always keep an Orval here if you visit Z{\"u}rich. Laurine, thanks for being there to share gossip at Montefiore. Guillaume Drion, Pierre Sacr{\'e}, and Laurent Mathys, thanks for our small chats during coffee breaks. Matthia, thanks for bringing some international flavours to the department and showing me some of your Italian cooking skills. I also want to thank all the others with whom I had the chance to spend lunch breaks or have small chats in the corridor: Anthony, Antonio, Anaïs, Jean-Michel, Gaspard, Balthazar, Yann, Marc, Philippe, Romain, Kathleen, Renaud, and Thibaud. Your friendship and support were greatly appreciated, and I feel lucky to have been part of such a wonderful group.

I am grateful to the ``Dream Team + Queens'', my friends outside of work, for being what they are. Andrea, Beniboy, Clairou, Laroushka, Mathilefou, Maximus, Pierro, Polo, R{\'e}my, QBask8, Thibaud, TZ: I am lucky to be part of such a beautiful group of friends. Spending time with you has helped me to maintain a healthy work-life balance and to have fun away from work-related things.

I want to express my heartfelt gratitude to my family. They have always been a source of love, encouragement, and inspiration. Their unwavering support and belief in me have been critical in helping me pursue my dreams. Thomas and mother, thanks for making fun of me and keeping me aware of the reality outside of academia. Marie, for always being my number 1 supporter. You are one of those who motivated me to pursue this PhD, and I will always be grateful for that. Dad, thank you for transferring to me your love for engineering and science and for always being present with good advice. I love you, dear family.

Finally, I would like to thank you, Alice, for your patience, love, and understanding throughout my PhD journey. Your unwavering support and encouragement have helped me overcome various challenges and motivated me to achieve my goals. I am deeply grateful to have you in my life.
% - Gilles could not proof read this part so I will be long and verbose. It is my way to show my sincerity.
%
% - I would like to thank the Jury member for accepting to judge the quality of my work.
%
% - Gilles.
%
% - The department.
%
% - My friends.
%
% - My family.
%
% - Alice.
